---
title: "Fantasy Football: Players Data"
author: "Odin Lee"
date: "Jan 26 2025"
format: "html"
---

# **Description**

Using **nflreadr** R package to create **PlayersData.csv** which will be used to import into the **Players Table** in the Fantasy Football Database.

## **Load libraries**

```{r}

# Load the needed libraries
library(tidyr)
library(dplyr)
library(stringr)
library(nflreadr)

```

## Import Data

```{r}

# Fetch player rosters for the 2021-2024 seasons
PlayersData <- load_rosters(2021:2024)

# Read in StatsData.csv file for later comparison
StatsData <- read.csv("StatsData.csv")

```

## Modify and Clean Data

```{r}

# Select relevant columns from PlayersData
PlayersDataSimplified <- PlayersData |>
  select(full_name, season, position, jersey_number, team, college, birth_date, height, weight,
         years_exp, headshot_url)

# Remove non QB, RB, WR, TE, and K
PlayersDataPlayers <- PlayersDataSimplified |>
  filter(position %in% c('QB', 'RB', 'WR', 'TE', 'K'))

# Separate full_name into to columns first_name and last_name
PlayersDataNames <- PlayersDataPlayers |>
  separate(full_name, into = c("first_name", "last_name"), sep = " ", extra = "merge")      
         
# Change team abbreviation for LA to LAR
MergedDataLA <- PlayersDataNames |>
  mutate(team = gsub("^LA$", "LAR", team))

```

## Normalize Names

This code normalizes names and ensures that only players being used are also in the **StatsData.csv** file.

```{r}

# Define a function to normalize names
NormalizeNames <- function(name) {
  name <- str_to_lower(name)  # Makes names lowercase
  name <- str_replace_all(name, "\\b(jr|sr|iii|ii|iv)\\b", "")  # Remove suffixes
  name <- str_replace_all(name, "\\.", "")  # Remove periods
  name <- str_squish(name)  # Trim excess spaces
  return(name)
}

# Apply the NormalizeNames function to first and last names
MergedDataLA$first_name <- NormalizeNames(MergedDataLA$first_name)
MergedDataLA$last_name <- NormalizeNames(MergedDataLA$last_name)

# Filter players to include only those present in the StatsData data set
PlayersDataCompared <- MergedDataLA |>
  semi_join(StatsData, by = c("first_name" = "first_name", "last_name" = "last_name"))

# Order PlayersDataCompared alphabeticaly by first_name then last_name then season
PlayersDataCompared <- PlayersDataCompared[order(PlayersDataCompared$first_name, 
                                                 PlayersDataCompared$last_name, 
                                                 PlayersDataCompared$season),]

```

## Clean birth_date Column

```{r}

# Replace NA values in the birth_date column for players who have a missing birth_date in one row but a valid birth_date in another
PlayersDataNAReplaced <- PlayersDataCompared |> 
  group_by(first_name, last_name) |>
  mutate(birth_date = if_else(is.na(birth_date), 
                              first(birth_date[!is.na(birth_date)]), 
                              birth_date)) |> 
  ungroup()

### NEXT CHUNK OF CODE IS TO VEIW THE DATA ###
# Check to see if there are any rows with NA values
PlayersWithNA <- PlayersDataNAReplaced |> 
  filter(if_any(everything(), is.na))

# Ensuring birth_date is formatted correctly
PlayersData <- PlayersDataNAReplaced |> 
  mutate(birth_date = as.Date(birth_date))

```

## Add Players Manually

```{r}

# Define additional players manually
additionalPlayers <- data.frame(
  first_name = c( "cedrick", "aj", "kavontae", "jeffery", "gabe", "avery", "robbie", 
                  "pj", "bam", "deonte", "pj", "robbie", "mike", "mike", 
                  "cameron", "dee", "dee", "dee", "joseph", "joseph", "joseph", 
                  "chig", "chig", "jeff", "jeff"
  ),
  last_name = c( "wilson", "dillon", "turpin", "wilson", "davis", "williams", "chosen",
                 "walker", "knight", "harty", "walker", "chosen", "badgley", "badgley",
                 "dicker", "eskridge", "eskridge", "eskridge", "fortson", "fortson", "fortson",
                 "okonkwo", "okonkwo", "wilson", "wilson"
  ),
  season = c( 2023, 2023, 2023, 2023, 2021, 2021, 2022, 
              2022, 2022, 2021, 2021, 2021, 2021, 2022, 
              2022, 2021, 2022, 2023, 2021, 2022, 2023, 
              2022, 2023, 2021, 2022
  ),
  position = c( "WR", "RB", "WR", "RB", "WR", "RB", "WR", 
                "QB", "RB", "WR", "QB", "WR", "K", "K", 
                "K", "WR", "WR", "WR", "TE", "TE", "TE", 
                "TE", "TE", "RB", "RB"
  ),
  jersey_number = c( 11, 28, 9, 23, 13, 35, 81, 
                     11, 27, 11, 11, 11, 6, 17, 
                     15, 1, 1, 1, 88, 88, 88, 
                     85, 85, 30, 23
  ),
  team = c( "MIA", "GB", "DAL", "MIA", "BUF", "ATL", "ARI", 
            "CAR", "NYJ", "NO", "CAR", "CAR", "TEN", "DET", 
            "LAC", "SEA", "SEA", "SEA", "KC", "KC", "KC", 
            "TEN", "TEN", "SF", "MIA"
  ),
  college = c( "Boise State", "Boston College", "Texas Christian", "North Texas", 
               "Central Florida", "Boise State", "Temple", "Temple", 
               "North Carolina State", "Assumption", "Temple", "Temple", 
               "Miami", "Miami", "Texas", "Western Michigan", 
               "Western Michigan", "Western Michigan", "Valdosta State", "Valdosta State",
               "Valdosta State", "Maryland", "Maryland", "North Texas", 
               "North Texas"
  ),
  birth_date = c( "1995-11-20", "1998-05-02", "1996-08-02", "1995-11-16", "1999-03-01", 
                  "1998-07-15", "1993-05-09", "1995-02-26", "2001-04-11", "1997-12-04", 
                  "1995-02-26", "1993-05-09", "1995-07-28", "1995-07-28", "2000-05-06", 
                  "1997-03-23", "1997-03-23", "1997-03-23", "1995-12-07", "1995-12-07", 
                  "1995-12-07", "1999-09-08", "1999-09-08", "1995-11-16", "1995-11-16"
  ),
  height = c( 74, 72, 69, 72, 74, 69, 75, 
              71, 70, 66, 71, 75, 70, 70, 
              73, 69, 69, 69, 78, 76, 76, 
              75, 75, 72, 72
  ),
  weight = c( 188, 250, 153, 194, 213, 195, 190, 
              212, 210, 170, 212, 190, 183, 183, 
              216, 190, 190, 190, 230, 230, 230, 
              250, 250, 194, 194
  ),
  years_exp = c( 5, 3, 4, 5, 4, 0, 6, 
                 5, 0, 2, 4, 5, 3, 4, 
                 0, 0, 1, 2, 2, 3, 4, 
                 0, 1, 3, 4
  ),
  headshot_url = c(
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/zvirmx1m9vycmfpfbf1w",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/dydcqdx4leta4rudagkq",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/dl2ngh2glo7gebocl3ah",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/rq7a1zluwjhoxrsuhpgi",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/gybdtwydpj0rggczumgg",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/kutrcxylkhvanrsbwhox",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/ymp4ipmdmmgi4omlw02f",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/cmuf0gtzectijxrktpg2",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/qdfkex1xfpfdyfovabig",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/ouuf2vot58iosguctjiy",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/cmuf0gtzectijxrktpg2",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/zgwiwdaqskofwe9rsvtj",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/vd3bpx6fhrx6thicas7c",
    "https://static.www.nfl.com/image/upload/f_auto,q_auto/league/ypn2qljwkyly4pdzt7vb",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/zv8qviwznwhdzeuxpni5",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/lwg4cpbjcdcgoru2emht",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/lwg4cpbjcdcgoru2emht",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/lwg4cpbjcdcgoru2emht",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/hmhdj4zf7rv2fiffzpg0",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/q4wdp6zldwvcte3dcbdp",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/gxhstzqswbn4z05r9rqv",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/sxuawjrgawwilhceso3d",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/hkqd6pubp52djhxhtyqi",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/zorqrtlrizlltgd6pnob",
    "https://static.www.nfl.com/image/private/f_auto,q_auto/league/zorqrtlrizlltgd6pnob"
  )
)

# Convert birth_date to Date in additional_players
additionalPlayers <- additionalPlayers |> 
  mutate(birth_date = as.Date(birth_date))

# Combine additional players with the main players data frame
PlayersData <- bind_rows(PlayersData, additionalPlayers)

```

## Export Data

```{r}

# Export the cleaned and filtered player data to a CSV file
write.csv(PlayersData, "PlayersData.csv", row.names = FALSE, na = "NULL")

```
